---
title: "Spectral signature"
format: html
editor: visual
---

---
title: "Spectral signature_all_dates"
output: html_document
author: "Polycarpe Mizero" 

format:
  html:
    code-fold: false
    embed-resources: true
    toc: true
    number-sections: false
    theme: cerulean
---

# seup

```{r warning=FALSE message=F }
library(tidyverse)
library(tidymodels)
library(ranger)
library(fs)
```

# Load and Combine All CSV Files

```{r}
# folder containing the CSV files
folder_path <- "../data/all_dates/"

# list csv files
csv_files <- dir_ls(folder_path, glob = "*.csv")

# read and combine
spectra_long <- map_dfr(csv_files, read_csv, show_col_types = FALSE)

head(spectra_long)

```
# EDA

```{r}
summary(spectra_long)
```

```{r}
glimpse(spectra_long)
```

# Pivot to Wide Format (Required for Random Forest)

Each wavelength becomes a predictor. Each row = one sample, each column = one reflectance feature. This massively reduces training time.

```{r warning=FALSE}
# Convert long -> wide format

spectra_wide <- spectra_long %>%
  select(date, sample_id, treatment, forage, rep, surface_type, wave_length, reflectance_percent) %>%
  pivot_wider(
    names_from = wave_length,
    values_from = reflectance_percent,
    names_prefix = "wl_"
  )%>%
  mutate(treatment = as.factor(treatment))

spectra_wide

```

## Ensure consistent factor levels before split

```{r}
# Make sure treatment is factor and has all levels
spectra_wide <- spectra_wide %>%
  mutate(treatment = factor(treatment)) %>%
  mutate(forage = factor(forage))%>%
  mutate(surface_type = factor(surface_type))%>%
  droplevels()

spectra_wide
```

#Last application date

```{r}
library(tidyverse)
library(lubridate)

app_dates <- as.Date(c("2025-06-12", "2025-08-12", "2025-09-15"))

spectra_wide <- spectra_wide %>%
  mutate(
    date = ymd(str_replace_all(date, "_", "-"))
  ) %>%
  arrange(date, sample_id) %>%
  mutate(
    last_app_date = sapply(date, function(d) max(app_dates[app_dates <= d])),
    DAA = as.numeric(date - last_app_date)
  )


spectra_wide
```

## application cycle to show where each observation belongs to (optional)

```{r}
spectra_wide <- spectra_wide %>%
  mutate(
    date = ymd(str_replace_all(date, "_", "-"))
  ) %>%
  arrange(date, sample_id) %>%
  mutate(
    app_cycle = case_when(
      date >= as.Date("2025-06-12") & date < as.Date("2025-08-12") ~ 1,
      date >= as.Date("2025-08-12") & date < as.Date("2025-09-15") ~ 2,
      date >= as.Date("2025-09-15") ~ 3,
      TRUE ~ NA_real_
    ),
    last_app_date = sapply(date, function(d) max(app_dates[app_dates <= d])),
    DAA = as.numeric(date - last_app_date)
  )

spectra_wide
```

# 5. RF CLASSIFICATION MODEL

#4 Spectral model with DAA added

Note: Random Forest predictors should only be spectral features.

```{r}
spectra_model <- spectra_wide %>%
  mutate(
    forage = as.factor(forage),
    treatment = as.factor(treatment),
    surface_type = as.factor(surface_type)
  ) %>%
  select(forage, treatment, DAA, starts_with("wl_"))

spectra_model
```

If you want to predict forage, this is fine because forage is the outcome and treatment, DAA, and wavelengths are predictors.

## For predicting forage, fold DAA into current workflow

```{r}

library(tidymodels)
library(ranger)

# Data for forage prediction
spectra_model_forage <- spectra_wide %>%
  mutate(
    forage = as.factor(forage),
    treatment = as.factor(treatment)
  ) %>%
  select(forage, treatment, DAA, starts_with("wl_"))

# Train/test split (70 / 30)

set.seed(123)
data_split <- initial_split(spectra_model_forage, prop = 0.7, strata = forage)

train_data <- training(data_split)
test_data  <- testing(data_split)

# Cross-validation
set.seed(123)
cv_folds <- vfold_cv(train_data, v = 5, strata = forage)

# Random Forest model
rf_model <- rand_forest(
  mode = "classification",
  mtry = tune(),
  trees = tune(),
  min_n = tune()
) %>%
  set_engine("ranger")

# Workflow
rf_workflow <- workflow() %>%
  add_model(rf_model) %>%
  add_formula(forage ~ .)

# Random tuning grid
rf_grid <- grid_random(
  mtry(range = c(10, 80)),
  trees(range = c(300, 800)),
  min_n(range = c(2, 10)),
  size = 20
)

# Tune
rf_tuned <- tune_grid(
  rf_workflow,
  resamples = cv_folds,
  grid = rf_grid,
  metrics = metric_set(accuracy, kap)
)

# Best parameters
best_params <- select_best(rf_tuned, metric = "accuracy")

# Final model
final_rf <- finalize_workflow(rf_workflow, best_params)

final_fit <- fit(final_rf, data = train_data)

# Predictions
predictions <- predict(final_fit, test_data) %>%
  bind_cols(test_data)

# Evaluation
accuracy(predictions, truth = forage, estimate = .pred_class)
conf_mat(predictions, truth = forage, estimate = .pred_class)

```

The formula forage \~ . already includes every other column as a predictor. So if DAA is in spectra_model, it is automatically included.

# Predicting surface type

## Option 1: Higher accuracy model

This includes forage, treatment, DAA, and spectra.

## model_surface_type

```{r}
spectra_model <- spectra_wide %>%
  mutate(
    forage = as.factor(forage),
    treatment = as.factor(treatment),
    surface_type = as.factor(surface_type)
  ) %>%
  select(surface_type, DAA, starts_with("wl_"))

spectra_model
```



```{r}
library(tidyverse)
library(tidymodels)
library(ranger)

tidymodels_prefer()

# --------------------------------------------------
# Data for surface_type prediction
# Higher-accuracy version: includes forage + treatment + DAA + spectra
# --------------------------------------------------
spectra_model_surface <- spectra_wide %>%
  mutate(
    date = as.character(date),
    sample_id = as.character(sample_id),
    forage = as.factor(forage),
    treatment = as.factor(treatment),
    surface_type = as.factor(surface_type),
    DAA = as.numeric(DAA)
  ) %>%
  drop_na() %>%
  select(date, sample_id, surface_type, forage, treatment, DAA, starts_with("wl_"))

# Optional: reduce spectral bands for speed
wl_cols <- names(spectra_model_surface)[str_detect(names(spectra_model_surface), "^wl_")]
wl_keep <- wl_cols[seq(1, length(wl_cols), by = 5)]

spectra_model_surface <- spectra_model_surface %>%
  select(date, sample_id, surface_type, forage, treatment, DAA, all_of(wl_keep))

# --------------------------------------------------
# Train/test split
# --------------------------------------------------
set.seed(123)
data_split <- initial_split(spectra_model_surface, prop = 0.7, strata = surface_type)

train_data <- training(data_split)
test_data  <- testing(data_split)

# --------------------------------------------------
# Cross-validation
# --------------------------------------------------
set.seed(123)
cv_folds <- vfold_cv(train_data, v = 3, strata = surface_type)

# --------------------------------------------------
# Recipe
# --------------------------------------------------
rf_recipe <- recipe(surface_type ~ ., data = train_data) %>%
  update_role(date, sample_id, new_role = "id") %>%
  step_zv(all_predictors())

# --------------------------------------------------
# Random Forest model
# --------------------------------------------------
rf_model <- rand_forest(
  mode = "classification",
  mtry = tune(),
  trees = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", importance = "impurity")

# --------------------------------------------------
# Workflow
# --------------------------------------------------
rf_workflow <- workflow() %>%
  add_recipe(rf_recipe) %>%
  add_model(rf_model)

# --------------------------------------------------
# Grid
# Adjust mtry so it does not exceed number of predictors
# --------------------------------------------------
n_pred <- train_data %>%
  select(-surface_type, -date, -sample_id) %>%
  ncol()

mtry_max_adj <- min(50, n_pred)
mtry_min_adj <- min(5, mtry_max_adj)

set.seed(123)
rf_grid <- grid_random(
  mtry(range = c(mtry_min_adj, mtry_max_adj)),
  trees(range = c(300, 800)),
  min_n(range = c(2, 10)),
  size = 20
)

# --------------------------------------------------
# Tune
# --------------------------------------------------
set.seed(123)
rf_tuned <- tune_grid(
  rf_workflow,
  resamples = cv_folds,
  grid = rf_grid,
  metrics = metric_set(accuracy, kap)
)

# --------------------------------------------------
# Best parameters
# --------------------------------------------------
best_params <- select_best(rf_tuned, metric = "accuracy")
best_params

# --------------------------------------------------
# Final model
# --------------------------------------------------
final_rf <- finalize_workflow(rf_workflow, best_params)

final_fit <- fit(final_rf, data = train_data)

# --------------------------------------------------
# Predictions
# --------------------------------------------------
predictions <- predict(final_fit, test_data) %>%
  bind_cols(test_data)

# --------------------------------------------------
# Evaluation
# --------------------------------------------------
surface_acc <- accuracy(predictions, truth = surface_type, estimate = .pred_class)
surface_cm  <- conf_mat(predictions, truth = surface_type, estimate = .pred_class)

surface_acc
surface_cm

# Numeric accuracy only
surface_acc$.estimate
```

## Option 2: Cleaner spectral-only model

This is would be a better choice if we want to say the model classified surface_type based mainly on hyperspectral response plus time since application.

```{r}
library(tidyverse)
library(tidymodels)
library(ranger)

tidymodels_prefer()

# --------------------------------------------------
# Data for surface_type prediction
# Cleaner version: DAA + spectra only
# --------------------------------------------------
spectra_model_surface <- spectra_wide %>%
  mutate(
    date = as.character(date),
    sample_id = as.character(sample_id),
    surface_type = as.factor(surface_type),
    DAA = as.numeric(DAA)
  ) %>%
  drop_na() %>%
  select(date, sample_id, surface_type, DAA, starts_with("wl_"))

# Optional: reduce spectral bands for speed
wl_cols <- names(spectra_model_surface)[str_detect(names(spectra_model_surface), "^wl_")]
wl_keep <- wl_cols[seq(1, length(wl_cols), by = 5)]

spectra_model_surface <- spectra_model_surface %>%
  select(date, sample_id, surface_type, DAA, all_of(wl_keep))

# --------------------------------------------------
# Train/test split
# --------------------------------------------------
set.seed(123)
data_split <- initial_split(spectra_model_surface, prop = 0.7, strata = surface_type)

train_data <- training(data_split)
test_data  <- testing(data_split)

# --------------------------------------------------
# Cross-validation
# --------------------------------------------------
set.seed(123)
cv_folds <- vfold_cv(train_data, v = 3, strata = surface_type)

# --------------------------------------------------
# Recipe
# --------------------------------------------------
rf_recipe <- recipe(surface_type ~ ., data = train_data) %>%
  update_role(date, sample_id, new_role = "id") %>%
  step_zv(all_predictors())

# --------------------------------------------------
# Random Forest model
# --------------------------------------------------
rf_model <- rand_forest(
  mode = "classification",
  mtry = tune(),
  trees = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", importance = "impurity")

# --------------------------------------------------
# Workflow
# --------------------------------------------------
rf_workflow <- workflow() %>%
  add_recipe(rf_recipe) %>%
  add_model(rf_model)

# --------------------------------------------------
# Grid
# --------------------------------------------------
n_pred <- train_data %>%
  select(-surface_type, -date, -sample_id) %>%
  ncol()

mtry_max_adj <- min(50, n_pred)
mtry_min_adj <- min(5, mtry_max_adj)

set.seed(123)
rf_grid <- grid_random(
  mtry(range = c(mtry_min_adj, mtry_max_adj)),
  trees(range = c(300, 800)),
  min_n(range = c(2, 10)),
  size = 20
)

# --------------------------------------------------
# Tune
# --------------------------------------------------
set.seed(123)
rf_tuned <- tune_grid(
  rf_workflow,
  resamples = cv_folds,
  grid = rf_grid,
  metrics = metric_set(accuracy, kap)
)

# --------------------------------------------------
# Best parameters
# --------------------------------------------------
best_params <- select_best(rf_tuned, metric = "accuracy")
best_params

# --------------------------------------------------
# Final model
# --------------------------------------------------
final_rf <- finalize_workflow(rf_workflow, best_params)

final_fit <- fit(final_rf, data = train_data)

# --------------------------------------------------
# Predictions
# --------------------------------------------------
predictions <- predict(final_fit, test_data) %>%
  bind_cols(test_data)

# --------------------------------------------------
# Evaluation
# --------------------------------------------------
surface_acc <- accuracy(predictions, truth = surface_type, estimate = .pred_class)
surface_cm  <- conf_mat(predictions, truth = surface_type, estimate = .pred_class)

surface_acc
surface_cm

# Numeric accuracy only
surface_acc$.estimate
```

# Predicting treatment (amendment)

## Model A: treatment without DAA

This tells how well spectra alone separate treatment classes.

```{r}
spectra_model_treatment_no_daa <- spectra_wide %>%
  mutate(
    treatment = as.factor(treatment),
    forage = as.factor(forage)
  ) %>%
  select(treatment, forage, starts_with("wl_"))
```

## Model B: treatment with DAA

This tells how much temporal information improves classification.

```{r}
spectra_model_treatment_with_daa <- spectra_wide %>%
  mutate(
    treatment = as.factor(treatment),
    forage = as.factor(forage)
  ) %>%
  select(treatment, forage, DAA, starts_with("wl_"))
```

Then use the same modeling workflow, just changing the outcome to treatment.

```{r}
# Train/test split
set.seed(123)
data_split <- initial_split(spectra_model_treatment_with_daa, prop = 0.7, strata = treatment)

train_data <- training(data_split)
test_data  <- testing(data_split)

# CV
set.seed(123)
cv_folds <- vfold_cv(train_data, v = 5, strata = treatment)

# RF model
rf_model <- rand_forest(
  mode = "classification",
  mtry = tune(),
  trees = tune(),
  min_n = tune()
) %>%
  set_engine("ranger")

# Workflow
rf_workflow <- workflow() %>%
  add_model(rf_model) %>%
  add_formula(treatment ~ .)

# Grid
rf_grid <- grid_random(
  mtry(range = c(10, 80)),
  trees(range = c(300, 800)),
  min_n(range = c(2, 10)),
  size = 20
)

# Tune
rf_tuned <- tune_grid(
  rf_workflow,
  resamples = cv_folds,
  grid = rf_grid,
  metrics = metric_set(accuracy, kap)
)

# Best params
best_params <- select_best(rf_tuned, metric = "accuracy")

# Final fit
final_rf <- finalize_workflow(rf_workflow, best_params)
final_fit <- fit(final_rf, data = train_data)

# Predict
predictions <- predict(final_fit, test_data) %>%
  bind_cols(test_data)

# Evaluate
accuracy(predictions, truth = treatment, estimate = .pred_class)
conf_mat(predictions, truth = treatment, estimate = .pred_class)
```



# Random Forest classification for all three targets:

surface_type

forage

treatment

## full script

```{r}
# =========================================================
# Random Forest workflow for hyperspectral classification
# Targets:
#   1. surface_type
#   2. forage
#   3. treatment (with and without DAA)
# =========================================================

# ---------------------------
# 1. Load libraries
# ---------------------------
#install.packages("doParallel")

library(tidyverse)
library(tidymodels)
library(ranger)
library(doParallel)
library(lubridate)

# Optional: faster / cleaner printing
tidymodels_prefer()

# ---------------------------
# 2. Prepare dataset
# ---------------------------
# Assumes spectra_wide already exists and contains:
# date, sample_id, treatment, forage, rep, surface_type, DAA, wl_350 ... wl_2500

# Make sure key columns are in the right format
spectra_wide <- spectra_wide %>%
  mutate(
    date = as.character(date),
    sample_id = as.character(sample_id),
    treatment = as.factor(treatment),
    forage = as.factor(forage),
    surface_type = as.factor(surface_type),
    rep = as.factor(rep),
    DAA = as.numeric(DAA)
  )

# Keep wavelength columns
wl_cols <- names(spectra_wide)[str_detect(names(spectra_wide), "^wl_")]

# Quick check
cat("Number of wavelength predictors:", length(wl_cols), "\n")

# ---------------------------
# 3. Optional: reduce bands for speed
# ---------------------------
# Keep every 5th wavelength to speed up tuning.
# Change by = 1 to keep all wavelengths.
wl_keep <- wl_cols[seq(1, length(wl_cols), by = 5)]

# Modeling dataset base
model_base <- spectra_wide %>%
  select(date, sample_id, treatment, forage, surface_type, DAA, all_of(wl_keep))

cat("Number of wavelength predictors retained:", length(wl_keep), "\n")

# ---------------------------
# 4. Start parallel processing
# ---------------------------
n_cores <- max(1, parallel::detectCores() - 1)
cl <- makePSOCKcluster(n_cores)
registerDoParallel(cl)

# ---------------------------
# 5. General modeling function
# ---------------------------
run_rf_model <- function(data, outcome, strata_var = NULL, seed = 123,
                         mtry_min = 5, mtry_max = 50,
                         trees_min = 300, trees_max = 800,
                         min_n_min = 2, min_n_max = 10,
                         grid_size = 20, v_folds = 3) {

  outcome <- rlang::ensym(outcome)

  # Remove rows with missing values
  data <- data %>%
    drop_na()

  # Define stratification column
  if (is.null(strata_var)) {
    strata_var <- rlang::as_name(outcome)
  }

  # Split
  set.seed(seed)
  data_split <- initial_split(
    data,
    prop = 0.7,
    strata = !!rlang::sym(strata_var)
  )

  train_data <- training(data_split)
  test_data  <- testing(data_split)

  # Cross-validation
  set.seed(seed)
  cv_folds <- vfold_cv(
    train_data,
    v = v_folds,
    strata = !!rlang::sym(strata_var)
  )

  # Recipe
  rf_recipe <- recipe(formula = as.formula(paste(rlang::as_name(outcome), "~ .")), data = train_data) %>%
    # Remove identifiers that should not be used as predictors
    update_role(date, sample_id, new_role = "id") %>%
    step_zv(all_predictors())

  # Model
  rf_model <- rand_forest(
    mode = "classification",
    mtry = tune(),
    trees = tune(),
    min_n = tune()
  ) %>%
    set_engine("ranger", importance = "impurity")

  # Workflow
  rf_workflow <- workflow() %>%
    add_recipe(rf_recipe) %>%
    add_model(rf_model)

  # Parameter range
  # mtry must not exceed number of predictors
  n_pred <- train_data %>%
    select(-all_of(rlang::as_name(outcome)), -date, -sample_id) %>%
    ncol()

  mtry_max_adj <- min(mtry_max, n_pred)
  mtry_min_adj <- min(mtry_min, mtry_max_adj)

  rf_grid <- grid_random(
    mtry(range = c(mtry_min_adj, mtry_max_adj)),
    trees(range = c(trees_min, trees_max)),
    min_n(range = c(min_n_min, min_n_max)),
    size = grid_size
  )

  # Tune
  set.seed(seed)
  rf_tuned <- tune_grid(
    rf_workflow,
    resamples = cv_folds,
    grid = rf_grid,
    metrics = metric_set(accuracy, kap)
  )

  # Best params
  best_params <- select_best(rf_tuned, metric = "accuracy")

  # Finalize
  final_rf <- finalize_workflow(rf_workflow, best_params)

  # Fit final model
  final_fit <- fit(final_rf, data = train_data)

  # Predict
  predictions <- predict(final_fit, test_data) %>%
    bind_cols(test_data)

  # Metrics
  acc <- accuracy(predictions, truth = !!outcome, estimate = .pred_class)
  kap_tbl <- kap(predictions, truth = !!outcome, estimate = .pred_class)
  cm <- conf_mat(predictions, truth = !!outcome, estimate = .pred_class)

  # Variable importance
  rf_engine <- extract_fit_parsnip(final_fit)$fit
  var_imp <- tibble(
    variable = names(rf_engine$variable.importance),
    importance = as.numeric(rf_engine$variable.importance)
  ) %>%
    arrange(desc(importance))

  # Return everything
  list(
    outcome = rlang::as_name(outcome),
    train_data = train_data,
    test_data = test_data,
    cv_folds = cv_folds,
    tuned_results = rf_tuned,
    best_params = best_params,
    final_workflow = final_rf,
    final_fit = final_fit,
    predictions = predictions,
    accuracy = acc,
    kappa = kap_tbl,
    confusion_matrix = cm,
    variable_importance = var_imp
  )
}

# ---------------------------
# 6. Build datasets for each target
# ---------------------------

# A. Surface type model
# Predict surface_type from DAA + treatment + forage + spectra
data_surface_type <- model_base %>%
  select(date, sample_id, surface_type, forage, treatment, DAA, all_of(wl_keep))

# B. Forage model
# Predict forage from DAA + treatment + spectra
data_forage <- model_base %>%
  select(date, sample_id, forage, treatment, DAA, all_of(wl_keep))

# C1. Treatment model without DAA
# Predict treatment from forage + spectra
data_treatment_no_daa <- model_base %>%
  select(date, sample_id, treatment, forage, all_of(wl_keep))

# C2. Treatment model with DAA
# Predict treatment from forage + DAA + spectra
data_treatment_with_daa <- model_base %>%
  select(date, sample_id, treatment, forage, DAA, all_of(wl_keep))

# ---------------------------
# 7. Run models
# ---------------------------

# Surface type classification
results_surface_type <- run_rf_model(
  data = data_surface_type,
  outcome = surface_type,
  strata_var = "surface_type",
  seed = 123,
  mtry_min = 5,
  mtry_max = 50,
  trees_min = 300,
  trees_max = 800,
  min_n_min = 2,
  min_n_max = 10,
  grid_size = 20,
  v_folds = 3
)

# Forage classification
results_forage <- run_rf_model(
  data = data_forage,
  outcome = forage,
  strata_var = "forage",
  seed = 123,
  mtry_min = 5,
  mtry_max = 50,
  trees_min = 300,
  trees_max = 800,
  min_n_min = 2,
  min_n_max = 10,
  grid_size = 20,
  v_folds = 3
)

# Treatment classification without DAA
results_treatment_no_daa <- run_rf_model(
  data = data_treatment_no_daa,
  outcome = treatment,
  strata_var = "treatment",
  seed = 123,
  mtry_min = 5,
  mtry_max = 50,
  trees_min = 300,
  trees_max = 800,
  min_n_min = 2,
  min_n_max = 10,
  grid_size = 20,
  v_folds = 3
)

# Treatment classification with DAA
results_treatment_with_daa <- run_rf_model(
  data = data_treatment_with_daa,
  outcome = treatment,
  strata_var = "treatment",
  seed = 123,
  mtry_min = 5,
  mtry_max = 50,
  trees_min = 300,
  trees_max = 800,
  min_n_min = 2,
  min_n_max = 10,
  grid_size = 20,
  v_folds = 3
)

# ---------------------------
# 8. Compare model performance
# ---------------------------

model_summary <- bind_rows(
  results_surface_type$accuracy %>%
    mutate(model = "surface_type"),
  results_forage$accuracy %>%
    mutate(model = "forage"),
  results_treatment_no_daa$accuracy %>%
    mutate(model = "treatment_no_DAA"),
  results_treatment_with_daa$accuracy %>%
    mutate(model = "treatment_with_DAA")
) %>%
  select(model, .estimate)

kappa_summary <- bind_rows(
  results_surface_type$kappa %>%
    mutate(model = "surface_type"),
  results_forage$kappa %>%
    mutate(model = "forage"),
  results_treatment_no_daa$kappa %>%
    mutate(model = "treatment_no_DAA"),
  results_treatment_with_daa$kappa %>%
    mutate(model = "treatment_with_DAA")
) %>%
  select(model, .estimate)

cat("\nAccuracy summary:\n")
print(model_summary)

cat("\nKappa summary:\n")
print(kappa_summary)

# ---------------------------
# 9. Print best parameters
# ---------------------------
cat("\nBest parameters - surface_type:\n")
print(results_surface_type$best_params)

cat("\nBest parameters - forage:\n")
print(results_forage$best_params)

cat("\nBest parameters - treatment without DAA:\n")
print(results_treatment_no_daa$best_params)

cat("\nBest parameters - treatment with DAA:\n")
print(results_treatment_with_daa$best_params)

# ---------------------------
# 10. Print confusion matrices
# ---------------------------
cat("\nConfusion matrix - surface_type:\n")
print(results_surface_type$confusion_matrix)

cat("\nConfusion matrix - forage:\n")
print(results_forage$confusion_matrix)

cat("\nConfusion matrix - treatment without DAA:\n")
print(results_treatment_no_daa$confusion_matrix)

cat("\nConfusion matrix - treatment with DAA:\n")
print(results_treatment_with_daa$confusion_matrix)

# ---------------------------
# 11. Save outputs if needed
# ---------------------------
# write_csv(model_summary, "rf_accuracy_summary.csv")
# write_csv(kappa_summary, "rf_kappa_summary.csv")
# write_csv(results_surface_type$variable_importance, "rf_varimp_surface_type.csv")
# write_csv(results_forage$variable_importance, "rf_varimp_forage.csv")
# write_csv(results_treatment_no_daa$variable_importance, "rf_varimp_treatment_no_daa.csv")
# write_csv(results_treatment_with_daa$variable_importance, "rf_varimp_treatment_with_daa.csv")

# ---------------------------
# 12. Stop parallel backend
# ---------------------------
stopCluster(cl)
registerDoSEQ()
```
























Three main figures 

1.  Spectral reflectance curves by class
2.  Random Forest variable importance plot
3.  Confusion matrix heatmap

# 1. Spectral reflectance curves by class

This figure shows the average reflectance pattern for each class across wavelengths.

```{r}
library(tidyverse)

# Mean reflectance by class and wavelength
spectral_summary <- spectra_long %>%
  group_by(treatment, wave_length) %>%
  summarise(
    mean_reflectance = mean(reflectance_percent, na.rm = TRUE),
    sd_reflectance = sd(reflectance_percent, na.rm = TRUE),
    .groups = "drop"
  )

# Plot
ggplot(spectral_summary, aes(x = wave_length, y = mean_reflectance, color = surface_type)) +
  geom_line(linewidth = 1) +
  geom_ribbon(
    aes(
      ymin = mean_reflectance - sd_reflectance,
      ymax = mean_reflectance + sd_reflectance,
      fill = surface_type
    ),
    alpha = 0.15,
    color = NA
  ) +
  labs(
    title = "Spectral Reflectance Curves by Surface Class",
    x = "Wavelength (nm)",
    y = "Reflectance (%)",
    color = "Surface type",
    fill = "Surface type"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

This figure shows:

Mean spectral signature of each class Separation among dung, urine, bare soil, and forage Variability around the mean using the shaded band



```{r}
ggsave("spectral_reflectance_curves.png", width = 10, height = 6, dpi = 300)

```

# 2. Random Forest variable importance plot

This identifies which wavelengths contributed most to classification.

If you used ranger through tidymodels, variable importance needs to be enabled in the model engine. If you have not done that yet, use this model specification:

```{r}
rf_model <- rand_forest(
  mode = "classification",
  mtry = tune(),
  trees = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", importance = "impurity")
```

Then refit the final model.

# Extract importance from the fitted model

```{r}
library(tidyverse)

# Extract fitted ranger model from workflow
rf_engine <- extract_fit_parsnip(final_fit)$fit

# Variable importance table
var_imp <- enframe(rf_engine$variable.importance, name = "wavelength", value = "importance") %>%
  arrange(desc(importance))
```

# Plot top important wavelengths

```{r}
var_imp %>%
  slice_max(order_by = importance, n = 20) %>%
  ggplot(aes(x = reorder(wavelength, importance), y = importance)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 20 Important Spectral Predictors in Random Forest Classification",
    x = "Wavelength",
    y = "Variable importance"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold")
  )
```

# If you want numeric wavelengths instead of wl_700 format

```{r}
var_imp <- var_imp %>%
  mutate(wavelength_num = readr::parse_number(wavelength))
```

Then plot:

```{r}
var_imp %>%
  slice_max(order_by = importance, n = 20) %>%
  ggplot(aes(x = reorder(wavelength_num, importance), y = importance)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 20 Important Wavelengths in Random Forest Classification",
    x = "Wavelength (nm)",
    y = "Variable importance"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold")
  )
```




```{r}
ggsave("variable_importance_plot.png", width = 8, height = 6, dpi = 300)

```

# 3. Confusion matrix heatmap



## Create confusion matrix data

```{r}
library(tidymodels)
library(tidyverse)

cm <- conf_mat(predictions, truth = surface_type, estimate = .pred_class)

cm_df <- as.data.frame(cm$table)
colnames(cm_df) <- c("Truth", "Prediction", "Freq")
```

## Plot confusion matrix heatmap

```{r}
ggplot(cm_df, aes(x = Prediction, y = Truth, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), size = 5) +
  labs(
    title = "Confusion Matrix Heatmap for Random Forest Classification",
    x = "Predicted class",
    y = "Reference class",
    fill = "Count"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Optional normalized confusion matrix

This version shows class-wise proportions.

```{r}
cm_df_prop <- cm_df %>%
  group_by(Truth) %>%
  mutate(Proportion = Freq / sum(Freq))

ggplot(cm_df_prop, aes(x = Prediction, y = Truth, fill = Proportion)) +
  geom_tile() +
  geom_text(aes(label = round(Proportion, 2)), size = 5) +
  labs(
    title = "Normalized Confusion Matrix for Random Forest Classification",
    x = "Predicted class",
    y = "Reference class",
    fill = "Proportion"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```



```{r}
ggsave("confusion_matrix_heatmap.png", width = 7, height = 6, dpi = 300)
```



